---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# SINT

SINT provides tools for specifying, analysing and simulating models of
social influence network theory based on the Friedkin-Johnsen model. The
latent dynamics

$$y(t+1) = \Lambda W y(t) + (I - \Lambda) y(0)$$

can be analysed in closed form (equilibrium, total influence matrix,
convergence diagnostics, also for signed networks) or simulated with influence
weights and susceptibilities that depend on time and on the state of the
system. An optional response stage maps latent opinions to manifest
responses, which can then be aggregated into collective outcomes.

## Installation

You can install the development version of SINT from GitHub with:

```r
# install.packages("remotes")
remotes::install_github("Silvestro26/SINT")
```

## Example

Three agents with influence matrix `W`, susceptibilities `lambda` and initial
opinions `y0`:

```{r example}
library(SINT)

W <- matrix(c(0.5, 0.5, 0,
              0.2, 0.6, 0.2,
              0,   0.3, 0.7), nrow = 3, byrow = TRUE)
lambda <- c(0.8, 0.5, 0.9)
y0 <- c(0, 0.5, 1)

fj_check(W, lambda)$converges
fj_equilibrium(W, lambda, y0)
```

The total influence matrix gives the weight of each initial opinion in each
equilibrium opinion:

```{r influence}
round(fj_influence(W, lambda), 3)
```

See `vignette("SINT")` for a complete guide to the package.
